---
title: "TreeSearch results"
author: "Martin R. Smith"
date: "`r Sys.Date()`"
---

# Parsimony analysis {#treesearch}

The phylogenetic dataset contains a high proportion of inapplicable codings (`r sum(my_chars == '-')`/`r length(my_chars)` = `r round(sum(my_chars == '-') / length(my_chars), 3) * 100`% of tokens), which are known to introduce error and bias to phylogenetic reconstruction when the Fitch algorithm is employed [@Maddison1993;@Brazeau2018].  As such, we employed a new tree-scoring algorithm that correctly handles inapplicable data [@Brazeau2018], implemented in the _MorphyLib_ _C_ library [@Brazeau2017Morphylib].  We employed the _R_ package _TreeSearch_ v0.1.2 [@Smith2018TreeSearch] to conduct phylogenetic tree search with this algorithm.

## Search parameters

Heuristic searches were conducted using the parsimony ratchet [@Nixon1999] under equal and implied weights [@Goloboff1997] with a variety of concavity constants.  The consensus tree presented in the main manuscript represents a strict consensus of all trees that are most parsimonious under one or more of the concavity constants (_k_) 2, 3, 4.5, 7, 10.5, 16 and 24, an approach that has been shown to produce higher accuracy than equal weights at any fixed level of precision [@Smith2017].

## Analysis 

### Load data

```{r treesearch-load-morphoBank, echo=FALSE}
# Load file from MorphoBank
my_data <- ReadAsPhyDat(filename)
my_data$Namacalathus <- NULL # Exclude Namacalathus

iw_data <- PrepareDataIW(my_data)
kValues <- c(2, 3, 4.5, 7, 10.5, 16, 24)
```

### Generate starting tree

_Dailyatia_ has been selected as an outgroup as camenellans have been interpreted as the earliest diverging members of the Brachiozoa [@Skovsted2015;@Zhao2017].

```{R treesearch-starting-tree}
nj.tree <- NJTree(my_data)
rooted.tree <- EnforceOutgroup(nj.tree, 'Dailyatia')
start.tree <- TreeSearch(tree=rooted.tree, dataset=my_data, maxIter=3000,
                         EdgeSwapper=RootedNNISwap, verbosity=0)
```

### Implied weights analysis

```{r treesearch-implied-weights-analysis, echo=TRUE, eval=FALSE}
for (k in kValues) {
  iw.tree <- IWRatchet(start.tree, iw_data, concavity=k,
                       ratchHits = 60, searchHits=55,
                       swappers=list(RootedTBRSwap, RootedSPRSwap, RootedNNISwap),
                       verbosity=0L)
  score <- IWScore(iw.tree, iw_data, concavity=k)
  # Write single best tree
  write.nexus(iw.tree,
              file=paste0("TreeSearch/hy_iw_k", k, "_", 
                          signif(score, 5), ".nex", collapse=''))

  suboptFraction = 0.02
  iw.consensus <- IWRatchetConsensus(iw.tree, iw_data, concavity=k,
                  swappers=list(RootedTBRSwap, RootedNNISwap),
                  searchHits=4,
                  suboptimal=score * suboptFraction,
                  nSearch=150, verbosity=0L)
  write.nexus(iw.consensus, 
              file=paste0("TreeSearch/hy_iw_k", k, "_", 
                          signif(IWScore(iw.tree, iw_data, concavity=k), 5),
                          ".all.nex", collapse=''))
}
```

### Equal weights analysis

```{r treesearch-equal-weights-analysis, echo=TRUE, eval=FALSE}
ew.tree <- Ratchet(start.tree, my_data, verbosity=0L,
                   ratchHits = 10, searchHits=55,
                   swappers=list(RootedTBRSwap, RootedSPRSwap, RootedNNISwap))
write.nexus(ew.tree, file=paste0(collapse='', "TreeSearch/hy_ew_",
                                 Fitch(ew.tree, my_data), ".nex"))

ew.consensus <- RatchetConsensus(ew.tree, my_data, nSearch=150,
                                 swappers=list(RootedTBRSwap, RootedNNISwap),
                                 verbosity=0L)
write.nexus(ew.consensus, file=paste0(collapse='', "TreeSearch/hy_ew_",
                                      Fitch(ew.tree, my_data), ".nex"))
```

## Results

### Implied weights results

```{R treesearch-load-IW-trees, echo=FALSE}
# Read results from files
iw.trees <- lapply(kValues, function (k) {
  iw.best <- list.files('TreeSearch', 
                      pattern=paste0('hy_iw_k',
                                     gsub('\\.', '\\\\.', k),
                                     '_\\d+\\.?\\d*\\.all\\.nex'),
                      full.names=TRUE)
  # Return:
  if (length(iw.best) == 0) {
    list()
  } else {
    read.nexus(iw.best[which.max(file.mtime(iw.best))])
  }
})
```



```{r treesearch-IW-consensus, fig.cap="Consensus of implied weights analyses at all values of k", fig.width=6, fig.height=5, echo=FALSE}
omit = c("Micrina", 'Mickwitzia_muralensis',
#omit=c(lon, 'Clupeafumosus_socialis',
       'Heliomedusa_orienta')
ColPlot(ConsensusWithout(lapply(iw.trees, consensus), omit))
ColMissing(omit)
```



```{r treesearch-all-IW-results, fig.cap="Implied weights results", fig.width=7.1, fig.height=10.6, echo=FALSE}
# Plot consensus results
par(mfrow=c(4, 2), mar=rep(0.2, 4))

ColPlot(consensus(lapply(iw.trees, consensus)))
text(-0.5, 1, pos=4, "Consensus of all k values", cex=0.8)

# Plot results for each value of k
for (i in seq_along(iw.trees)) {
  ColPlot(consensus(iw.trees[[i]]))
  text(1, 1, paste0('k = ', kValues[i]), pos=4)
}

```

### Equal weights results

```{r treesearch-equal-weights-results, fig.cap="Strict consensus of equal weights results", echo=FALSE}
ew.best <- list.files('TreeSearch', pattern='hy_ew_\\d*\\.nex', full.names=TRUE)
ew.trees <- read.nexus(file=ew.best[which.max(file.mtime(ew.best))])
ColPlot(consensus(ew.trees))
```

```{R treesearch-equal-weights-pruned-consensus, fig.cap="Strict consensus of equal weights results, taxa excluded", echo=FALSE, eval=FALSE}
# omit <- c("none")
# ColPlot(ConsensusWithout(ew.tree, omit))
# ColMissing(omit)
```
